---
title: "Tidy Tuesday Week 6"
description: |
  My spatial exploration of the global plastics pollution TidyTuesday dataset.
author:
  - name: Alex Ehrens
date: 02-25-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# attach packages
library(tidyverse)
library(sf)
library(tidytuesdayR)
library(ggspatial)
library(maptools)
library(ghibli)
library(wesanderson)
```

I would like to present my submission for Tidy Tuesday (from 1/26), which was a dataset of global plastics pollution. I used the maptools package to get the "wrld_simpl" dataset with geometries for every country and then used sf to transform it into a spatial dataframe. I then joined this to the plastic pollution dataset by country name using [this method](https://github.com/Z3tt/TidyTuesday/blob/master/R/2019_21_PlasticWaste.Rmd) from another Tidy Tuesday coder I noticed that I had to manually change some of the country names (like the U.S.) so that the datasets joined properly. Then for the coloring, I used the `{wesanderson}` package by [karthik](https://github.com/karthik/wesanderson) and found a color palette from The Life Aquatic with Steve Zissou. I didn't like how the countries without plastics data wouldn't show up on the map so I plotted the `wrld_simpl` country polygons in the background to fill out the map. I'm very glad with how this turned out!

### Read in Tidy Tuesday plastics data
```{r, message=FALSE}
tuesdata <- tidytuesdayR::tt_load('2021-01-26')

plastics <- tuesdata$plastics
```

### Select and filter plastics data to just get totals
```{r}
plastics_totals <- plastics %>% 
  select(country, parent_company, grand_total, volunteers) %>% 
  filter(parent_company == "Grand Total") %>% 
  filter(country != "EMPTY") %>% 
  filter(country != "Taiwan_ Republic of China (ROC)") %>% 
  filter(country != "NIGERIA") %>% 
  filter(country != "Philippines") # these 3 countries seem to be an outlier that is way too high

plastics_totals$country[plastics_totals$country == "ECUADOR"] <- "Ecuador"
plastics_totals$country[plastics_totals$country == "United States of America"] <- "United States"
plastics_totals$country[plastics_totals$country == "Cote D_ivoire"] <- "Cote d'Ivoire"
plastics_totals$country[plastics_totals$country == "Vietnam"] <- "Viet Nam"
#plastics_totals$country[plastics_totals$country == "NIGERIA"] <- "Nigeria"
plastics_totals$country[plastics_totals$country == "Tanzania"] <- "United Republic of Tanzania"
#plastics_totals$country[plastics_totals$country == "Taiwan_ Republic of China (ROC)"] <- "Taiwan"
```

### Read in world polygons data
```{r}
data("wrld_simpl")
```

### Join plastics data with world spatial dataset
```{r}
plastics_totals_spatial <- wrld_simpl %>% 
  st_as_sf() %>% 
  st_transform(crs = 4326) %>% 
  mutate(NAME = as.character(NAME)) %>% 
  left_join(plastics_totals, by = c("NAME" = "country")) %>% 
  filter(is.na(grand_total) == FALSE)

# make sure coordinate reference systems match
world_total <- wrld_simpl %>% 
  st_as_sf() %>% 
  st_transform(crs = 4326)
```

### Map total plastics production by country
```{r}
palette <- wes_palette("Zissou1", type = "continuous")

ggplot() +
  geom_sf(data = world_total,
          aes(geometry = geometry)) +
  geom_sf(data = plastics_totals_spatial,
          aes(geometry = geometry, fill = grand_total)) +
  scale_fill_gradientn(colors = palette) +
  theme_minimal() +
  labs(title = "Global plastics pollution as told by The Life Aquatic",
       fill = "Plastic pollution \n(counts)") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.key.width = unit(2, "cm"))
```

